ComBatLS LBCC Results

### 2025-01-23 eda/lifespan_eda.Rmd
remaking LBCC plots to include country

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(gamlss, gamlss.cens, gamlss.mx, gamlss.tr, ggplot2, tidyverse, ggpubr, skimr, gamlss.data, data.table, nlme, devtools, slider, see, broom.mixed)

@@

Load data from Jakob - updated with euler numbers and corrected VETSA study
```{r}
cn.df <- fread(paste0(base_path,"/data/v1_2_1_nature.csv"), stringsAsFactors = TRUE)
```

@@

```{r}
#clean up site NAs
cn.df.imp1 <- cn.df.eulerfilt2 %>%
  mutate(site = trimws(as.character(site)),
         study = trimws(study)) %>%
  mutate(site = gsub(" ", "_", ifelse(is.na(site), study, paste(study, site, sep="_")))) 
skim(cn.df.imp1)
```

```{r}
#drop NAs listwise
cn.df.imp <- cn.df.imp1%>%
  dplyr::select(!c(INDEX.ID, country)) %>%#drop index ID for filtering
  na.omit()

@@

Demographics

@@

#save names with site (since not saved correctly from centile outputs)
study_site_key <- cn.df.imp %>%
  dplyr::select(study, site) %>%
  unique()
n_unique(study_site_key$study) == n_unique(cn.df.imp$study)
n_unique(study_site_key$site) == n_unique(cn.df.imp$site)
fwrite(study_site_key, file=paste0(base_path, "/data/lifespan_study_site_key.csv"))
```

>>>>

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(gamlss, gamlss.cens, gamlss.mx, gamlss.tr, ggplot2, tidyverse, ggpubr, skimr, gamlss.data, data.table, nlme, devtools, slider, see, broom.mixed, ggrain, ggpubr)

@@

Load data from Jakob - updated with euler numbers and corrected VETSA study
```{r}
cn.df <- fread(paste0(base_path,"/data/v1_2_1_nature.csv"), stringsAsFactors = TRUE, na.strings="")
```

@@

```{r}
#clean up site NAs
cn.df.imp1 <- cn.df.eulerfilt2 %>%
  mutate(site = gsub(" ", "_", trimws(as.character(site))),
         study = gsub(" ", "_", trimws(as.character(study)))) %>%
  mutate(site = ifelse(is.na(site), study, paste(study, site, sep="_"))) 
skim(cn.df.imp1)
```

```{r}
#drop NAs listwise
cn.df.imp <- cn.df.imp1%>%
  dplyr::select(!c(INDEX.ID)) %>%#drop index ID for filtering
  na.omit()

@@

##Demographics

@@

#save names with site (since not saved correctly from centile outputs)
study_site_key <- cn.df.imp %>%
  dplyr::select(study, site, country, fs_version) %>%
  unique()
n_unique(study_site_key$study) == n_unique(cn.df.imp$study)
n_unique(study_site_key$site) == n_unique(cn.df.imp$site)
fwrite(study_site_key, file=paste0(base_path, "/data/lifespan_study_site_key.csv"))
```

###Plot

Clean up names, order by age
```{r}
#reorder studies by age
age.df <- cn.df.imp %>%
  group_by(study) %>%
  summarize(mean_age = mean(age_days)) %>%
  arrange(mean_age)
study_names_clean <- gsub("_", " ", unique(age.df$study))

final_df <- cn.df.imp %>% 
  mutate(age_years=(age_days-280)/365.25,
         study = gsub("_", " ", study)) %>%
  mutate(study = factor(study, levels=study_names_clean, ordered=TRUE))
```

Sample plot
```{r}
sample_boxplot <- final_df %>%
  #mutate(study = factor(study, levels=c(age.df$study), ordered=TRUE)) %>%
ggplot(aes(y=study, x=age_years)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  #guides(y = ggh4x::guide_axis_nested(delim = "&")) +
  theme(axis.title.x=element_blank(), axis.ticks.y = element_blank(), axis.text.x=element_blank()) +
  labs(color="Sex", y = "Study")

#overall sample
full_boxplot <- final_df %>%
  mutate(ylab = "   Full Sample") %>%
  ggplot(aes(y=ylab, x=age_years, color=sex)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  # theme(plot.margin = margin(0,0,0,0, 'cm')) +
  # scale_y_discrete(labels=c("1" = "Full Sample")) +
  labs(color="Sex", x = "Age (Years)", y="")

sample_barplot <- final_df %>%
  #mutate(study = factor(study, levels=c(age.df$study), ordered=TRUE)) %>%
  ggplot() +
  geom_bar(aes( x=..count../tapply(..count.., ..y.. ,sum)[..y..], y=study, fill=sex)) +
  scale_x_continuous(labels = scales::percent) +
  geom_vline(xintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  theme(axis.ticks.y = element_blank(), axis.text=element_blank(), axis.title=element_blank()) +
  labs(fill="Sex")

full_barplot <- final_df %>%
  mutate(ylab = "Full Sample") %>%
  ggplot() +
  geom_bar(aes(x=..count../tapply(..count.., ..y.. ,sum)[..y..], y=ylab, fill=sex)) + 
  scale_x_continuous(labels = scales::percent) +
  geom_vline(xintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  theme(axis.ticks.y = element_blank(), axis.text.y=element_blank(), axis.title.y=element_blank()) +
  labs(fill="Sex", x = "Sex")

ggarrange(sample_boxplot, sample_barplot,
          full_boxplot, full_barplot,
          nrow=2, ncol=2, 
          align = "hv", 
          common.legend = TRUE,
          widths=c(2, 1),
          heights=c(10, 1))
```

look at country info

```{r}
sample_boxplot2 <- final_df %>%
  ggplot(aes(y=age_years, x=study, fill=country)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  #guides(y = ggh4x::guide_axis_nested(delim = "&")) +
  theme(axis.title.x=element_blank(), axis.text.x = element_blank(), legend.position = "right") +
  labs(fill="Country", y = "Age (Years)")  +
  paletteer::scale_fill_paletteer_d("ggthemes::Tableau_20")
```

```{r}
blank <- ggplot() + theme_void()

#overall sample
full_boxplot2 <- final_df %>%
  mutate(xlab = "Full Sample ") %>%
  ggplot(aes(x=xlab, y=age_years, color=sex)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  theme(axis.title=element_blank(), axis.text=element_blank(), axis.ticks.y=element_blank()) +
  labs(color="Sex", x=" ")

sample_barplot2 <- final_df %>%
  ggplot() +
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], x=study, fill=sex)) +
  geom_hline(yintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_text(angle=90, hjust=0.5), axis.text.y = element_text(size=6)) +
  labs(fill="Sex", x = "Study", y="Percent Sex")

full_barplot2 <- 
  final_df %>%
  mutate(xlab = "  Full Sample  ") %>%
  ggplot() +
  geom_bar(aes(y=..count../tapply(..count.., ..x.. ,sum)[..x..], x=xlab, fill=sex)) + 
  geom_hline(yintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  theme(axis.ticks.y = element_blank(), axis.title=element_blank(), axis.text.y=element_blank(), axis.text.x = element_text(angle=90, hjust=0.5, size=12)) +
  labs(color="Sex", y = "Age (Years)", y="")

full_rain <- final_df %>%
  ggplot(aes(x=sex, y=age_years, fill=sex)) +
  geom_rain(alpha=.4) +
  theme_linedraw(base_size = 11) +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_text(angle=90, hjust=0.5), 
        axis.title.x = element_blank(), axis.text.y = element_text(size=6),
        legend.position = "none") +
  labs(y="Age (Years)")
```

```{r}
country_leg <- as_ggplot(get_legend(sample_boxplot2))
sex_leg <- as_ggplot(get_legend(sample_barplot2))

fig5A <- ggarrange(sample_boxplot2, country_leg,
          sample_barplot2,  sex_leg,
          nrow=2, ncol=2, 
          #align = "v", 
          common.legend = FALSE,
          legend = "none",
          widths=c(11, 1),
          heights=c(1.9, 1)
          )
```

```{r}
fig5_AB <- ggarrange(fig5A, full_rain,
          nrow=1, ncol=2, 
          legend = "none",
          widths=c(3, 1),
          labels = "AUTO")

saveRDS(fig5_AB, file="../figs/figure_5_AB.rds")
```

### 2025-01-23 paper_lbcc_results.Rmd
remaking LBCC plots to include country

gam_centile_df <- full_join(gam_centile_df, study_site_key) %>% 
  mutate(age_years=age_days/365.25)
n_unique(gam_centile_df$study)

dim(gam_centile_df)
```

Combat-LS Centiles
```{r}
gamlss_centile_df <- fread(file=paste0(der_path, "/lifespan-CN-imp-sites-euler-log-cf-cf.gamlss-batch.study_predictions.csv"))
gamlss_centile_df <- full_join(gamlss_centile_df, study_site_key) %>% 
  mutate(age_years=age_days/365.25)

@@

Unharmonized Centiles
```{r}
raw_centile_df <- fread(file=paste0(der_path, "/lifespan-CN-imp-sites-euler-raw_predictions.csv"))
raw_centile_df <- full_join(raw_centile_df, study_site_key) %>% 
  mutate(age_years=age_days/365.25)

@@

```{r}
deltas_df <- fread(file=paste0(der_path, "/batch.study_pred_deltas.csv"))
deltas_df <- full_join(deltas_df, study_site_key) %>% 
  mutate(age_years=age_days/365.25)

@@

Sample plot
```{r}
#reorder studies by age
age.df <- gamlss_centile_df %>%
  group_by(study) %>%
  summarize(mean_age = mean(age_years)) %>%
  arrange(mean_age)

sample_boxplot <- gamlss_centile_df %>%
  mutate(study = factor(study, levels=c(age.df$study), ordered=TRUE)) %>%
ggplot(aes(y=study, x=age_years)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  #guides(y = ggh4x::guide_axis_nested(delim = "&")) +
  theme(axis.title.x=element_blank(), axis.ticks.y = element_blank(), axis.text.x=element_blank()) +
  labs(color="Sex", y = "Study")

#overall sample
full_boxplot <- 
  gamlss_centile_df %>%
  mutate(ylab = "   Full Sample") %>%
  ggplot(aes(y=ylab, x=age_years, color=sex)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  # theme(plot.margin = margin(0,0,0,0, 'cm')) +
  # scale_y_discrete(labels=c("1" = "Full Sample")) +
  labs(color="Sex", x = "Age (Years)", y="")

sample_barplot <- gamlss_centile_df %>%
  mutate(study = factor(study, levels=c(age.df$study), ordered=TRUE)) %>%
  ggplot() +
  geom_bar(aes( x=..count../tapply(..count.., ..y.. ,sum)[..y..], y=study, fill=sex)) +
  scale_x_continuous(labels = scales::percent) +
  geom_vline(xintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  theme(axis.ticks.y = element_blank(), axis.text=element_blank(), axis.title=element_blank()) +
  labs(fill="Sex")

full_barplot <- 
  gamlss_centile_df %>%
  mutate(ylab = "Full Sample") %>%
  ggplot() +
  geom_bar(aes(x=..count../tapply(..count.., ..y.. ,sum)[..y..], y=ylab, fill=sex)) + 
  scale_x_continuous(labels = scales::percent) +
  geom_vline(xintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  theme(axis.ticks.y = element_blank(), axis.text.y=element_blank(), axis.title.y=element_blank()) +
  labs(fill="Sex", x = "Sex")

ggarrange(sample_boxplot, sample_barplot,
          full_boxplot, full_barplot,
          nrow=2, ncol=2, 
          align = "hv", 
          common.legend = TRUE,
          widths=c(2, 1),
          heights=c(10, 1))
```

@@

```{r}
blank <- ggplot() + theme_void()

sample_boxplot2 <- gamlss_centile_df %>%
  mutate(study = factor(study, levels=c(age.df$study), ordered=TRUE)) %>%
ggplot(aes(y=age_years, x=study, color=sex)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  #guides(y = ggh4x::guide_axis_nested(delim = "&")) +
  theme(axis.title.x=element_blank(), axis.text.x = element_blank()) +
  labs(color="Sex", y = "Age\n(Years Post-Conception)")

#overall sample
full_boxplot2 <- 
  gamlss_centile_df %>%
  mutate(xlab = "Full Sample ") %>%
  ggplot(aes(x=xlab, y=age_years, color=sex)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  theme(axis.title=element_blank(), axis.text=element_blank(), axis.ticks.y=element_blank()) +
  # scale_y_discrete(labels=c("1" = "Full Sample")) +
  labs(color="Sex", x=" ")

sample_barplot2 <- gamlss_centile_df %>%
  mutate(study = factor(study, levels=c(age.df$study), ordered=TRUE)) %>%
  ggplot() +
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], x=study, fill=sex)) +
  #scale_y_continuous(labels = scales::percent) +
  geom_hline(yintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  scale_y_continuous(labels = scales::percent) +
  #theme(axis.ticks.x = element_blank(), axis.text=element_blank(), axis.title=element_blank()) +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_text(angle=90, hjust=0.5), axis.text.y = element_text(size=6)) +
  labs(fill="Sex", x = "Study", y="Percent Sex")

full_barplot2 <- 
  gamlss_centile_df %>%
  mutate(xlab = "  Full Sample  ") %>%
  ggplot() +
  geom_bar(aes(y=..count../tapply(..count.., ..x.. ,sum)[..x..], x=xlab, fill=sex)) + 
  geom_hline(yintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  theme(axis.ticks.y = element_blank(), axis.title=element_blank(), axis.text.y=element_blank(), axis.text.x = element_text(angle=90, hjust=0.5, size=12)) +
  labs(color="Sex", y = "Age\n(Years Post-Conception)", y="")
```

>>>>

gam_centile_df <- left_join(gam_centile_df, study_site_key) %>% 
  mutate(age_years=(age_days-280)/365.25)
n_unique(gam_centile_df$study)

dim(gam_centile_df)

gam_centile_df %>% dplyr::filter(is.na(age_days))
```

Combat-LS Centiles
```{r}
gamlss_centile_df <- fread(file=paste0(der_path, "/lifespan-CN-imp-sites-euler-log-cf-cf.gamlss-batch.study_predictions.csv"))
gamlss_centile_df <- full_join(gamlss_centile_df, study_site_key) %>% 
    mutate(age_years=(age_days-280)/365.25)

@@

Unharmonized Centiles
```{r}
raw_centile_df <- fread(file=paste0(der_path, "/lifespan-CN-imp-sites-euler-raw_predictions.csv"))
raw_centile_df <- full_join(raw_centile_df, study_site_key) %>% 
    mutate(age_years=(age_days-280)/365.25)

@@

```{r}
deltas_df <- fread(file=paste0(der_path, "/batch.study_pred_deltas.csv"))
deltas_df <- full_join(deltas_df, study_site_key) %>% 
    mutate(age_years=(age_days-280)/365.25)

@@

@@
